---
title: "Implications of Sex Ed on Teen Sexual Health by State"
author: "Aditi Nayak, Lauren Pelosi, Catherine Sarosi"
date: "11/19/2020"
output:
  rmdformats::readthedown:
    thumbnails: false
    highlight: NULL
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, fig.align="center")
library(tidyverse)
library(shiny)
```

# Introduction

<div style="width:100%;height:0;padding-bottom:56%;position:relative;"><iframe src="https://giphy.com/embed/3o7aTDP5aDFogC5KmY" width="100%" height="100%" style="position:absolute" frameBorder="0" class="giphy-embed" allowFullScreen></iframe></div><p><a href="https://giphy.com/gifs/filmeditor-movie-mean-girls-3o7aTDP5aDFogC5KmY">via GIPHY</a></p>

Sexual health is still a taboo topic across the United States. The openness in conversations about this topic vary by state, community, family, and even individual. As a result, students across the nation have vastly different sexual education experiences: some like the above scene in Mean Girls and others more comprehensive. 

This inconsistency and inequity in sexual education requirements across the nation is not only awkward but can also be consequential for teens. Like an accounting course prepares students for future finances and English classes helps students communicate better, sexual education helps students navigate their sexuality and safely engage in sexual activity. Without sexual education, students are less prepared to manage their sexual and reproductive health and behavior, which can lead to unwanted pregnancy and sexually transmitted diseases (STD).

Given how consequential sexual education can be to teenagers, we investigated how sexual education impacts certain health outcomes. The general effectiveness of sexual education is a broad topic, so we narrowed in on the following questions:

## Research Questions
(1) How do state sexual health education mandates affect student health outcomes? 
(2) Do state education mandates correspond to the policies schools enact and the material teachers choose to emphasize?
(3) How do school policies and teachers’ choices affect student health outcomes? 
(4) What groupings form based on health outcomes? 

## Data

To answer these questions, we collected the following data:

  - **Statewide Education Mandates**: To assess how effective statewide sexual education mandates are, we need to know what each state requires as a minimum of all high school sex ed programs. We collected this data from a [Guttmacher Institute](https://www.guttmacher.org/state-policy/explore/sex-and-hiv-education#) survey last updated November 1, 2020, which noted if each state requires of sex ed programs to have certain characteristics.


  - **Percent of High Schools in State with Certain Sex Ed Programs**: Even with statewide sex ed mandates, not all schools within the same state will offer similar sex ed programs.To examine what a state’s schools actually do, we collected data from the [CDC](https://www.cdc.gov/healthyyouth/policy/state_law_summary_reports.htm). In 2017, the CDC surveyed high schools across the nation about what their school’s sex ed curriculum entailed. The CDC then reported what percent of high schools in a state teach certain topics.
    
    
  - **Sexual Behaviors**: One way we assess the effectiveness of these sex ed programs is through what students acutally do. The aforementioned [CDC 2017 survey report](https://www.cdc.gov/healthyyouth/policy/state_law_summary_reports.htm) offered us this information (such as what percent of students in a state have had sexual intercourse with at least 4 partners). 
	
	
  - **Health Outcomes**: We also looked at related student health outcomes (such as chlamydia cases and number of teenage pregnancies) to measure the effectiveness of sex ed programs. We collected this information from the Centers for Disease Control and Prevention’s 2018 STD Surveillance Report, as reported by [Alarms.org](https://www.alarms.org/std-statistics/). We combined information from Alarms.org with data from [Power to Decide](https://powertodecide.org/what-we-do/information/national-state-data/teen-birth-rate) about teen birth rate among girls ages 15-19.

```{r tableOutcomes, echo=FALSE, message=FALSE, warnings=FALSE, results='asis'}
tablOutcomes <- "
| Variable        | Definition           | 
|---------------|:-------------:|
| Sex Ed Mandated  |  Sex education is mandated in some capacity in the state |
| HIV Ed Mandated | HIV education is mandated in some capacity in the state  |
| Age Appropriate | When provided, sex or HIV education must be age-appropriate      |
| Culturally Appropriate/Unbiased  |  When provided, sex or HIV education must be culturally appropriate/unbiased |
| Cannot Promote Religion | When provided, sex or HIV education cannot promote religion   |
| Parents Notified | ...that sex education is occurring   |
| Parental Consent | ...is required for sex education  |
| Opt-Out | Parents can choose to opt-out of sex education for their child |
| Sex Ed Contraception | When provided, sex education must cover contraception  |
| Sex Ed Abstinence | When provided, sex education must emphasize abstinence  |
| Sex Ed Marriage | When provided, sex education must emphasize the importance of sex only within marriage |
| Sex Ed Orientation | When provided, sex education must cover sexual orientation  |
| Sex Ed Negative Outcomes | When provided, sex education must cover the possible negative outcomes of sex |
| HIV Ed Condoms | When provided, HIV education must provide condoms |
| HIV Ed Abstinence | When provided, HIV education cover or stress abstinence |
"
cat(tablOutcomes) # output the table in a format good for HTML/PDF/docx conversion
```



More information about what the data represents or how it was collected by these sources can be found through the above hyperlinks. Further information about the data used can also be found on the following table:

# How do state sexual health education mandates affect student health outcomes? 

## Are stricter mandates effective in improving student health outcomes?

Here, we explore the relationship between statewide health education mandates and metrics of student health and wellbeing. 

<iframe height="850" width="100%" frameborder="no" title="Histogram1" src="https://anayak23.shinyapps.io/Blog_L1/"> </iframe>

```{r table2, echo=FALSE, message=FALSE, warnings=FALSE, results='asis'}
tabl <- 
"| Education Mandate        | Health Outcome  |  Interesting Insight |
|---------------|:-------------:|:-------------:|
| Sexual decision making and self-discipline | % students intoxicated | Suggests that education on sexual decision making may not be effective, since those states where most students were intoxicated during their last sexual encounter do mandate this type of education. |
| Sex Ed Marriage | % students with one or more partners  | The three states in which the most students have had four or more sexual partners also mandate sex ed that emphasises waiting for marriage to have sex. |"
cat(tabl) # output the table in a format good for HTML/PDF/docx conversion
```


There seems to be a disconnect between the severity of a state’s education mandates and its related student health outcomes. Of course, lurking variables are at play. Other variables besides health education correlate with geographic location and affect student health; these include environmental factors, food, alcohol, and tobacco culture, and access to outdoor space for exercise. However, it seems to us that schools’ empowering students to make informed choices likely does have some effect on their wellbeing, contrary to what this visualization suggests. Therefore, we were motivated to look next at education practices - what actually happens inside schools - to see if they align with mandates. If practices and mandates do not coincide, that may explain why we don’t see a clear effect of mandates on student health outcomes. 

# Do state education mandates correspond to the policies schools enact and the material teachers choose to emphasize?

## How do statewide mandates and that state’s high schools’ sex ed programs compare?

States mandate certain minimum requirements for secondary schools’ sex ed programs; however, schools can decide how they will comply with the mandate and whether they will enact policies that go beyond the mandate. Further, teachers have agency over how they will teach the material. Here we investigate the overlap between mandates and the programs that educators put into practice. 

<iframe height="850" width="100%" frameborder="no" title="ScatterEdu" src="https://anayak23.shinyapps.io/edu_scatter/"> </iframe>

```{r tableA2, echo=FALSE, message=FALSE, warnings=FALSE, results='asis'}
##
tablA2 <- "
| Education Mandate        | Resource Offered by High Schools  |  Interesting Insight |
|---------------|:-------------:|:-------------:|
| HIV Ed Mandated | Increased Student Knowledge on HIV Prevention  | There are 8 states in which sex ed is mandated but in which fewer than 70% of schools provide teachers strategies for teaching sex ed. In these states, teachers have a large amount of discretion. (However, it is unclear if this disparity between mandate and policy has an effect on student health. All of these 8 states have an STD index of 0.45 or lower, scoring better than about half of all states. In fact, one of the states, Vermont, has the lowest STD index of any state. By another metric, births per 1,000 girls, this result is less definitive: the states fall over a wide range, from 27.2 (Kentucky) to 8.8 (Vermont).) |
| Sex Ed Orientation | Increased Student Knowledge on Sexuality | Without a mandate (coded as 0) and with a mandate (coded as 1), the percentage of teachers that tried to increase knowledge on human sexuality is highly concentrated from 70-90%. Due to the high variance in the percentage among states without a mandate, the average percentage is lower than in states with a mandate. |"
cat(tablA2) # output the table in a format good for HTML/PDF/docx conversion
```

We see some preliminary evidence that mandates have a small, but not definitive, effect on teacher practices. 

<iframe height="1000" width="100%" frameborder="no" title="Histogram3" src="https://anayak23.shinyapps.io/Blog_L3/ "> </iframe>

```{r tableLP, echo=FALSE, message=FALSE, warnings=FALSE, results='asis'}
##Corresponds with lauren_blog3
tablLP <- "
| Education Mandate        | Practice/Behavior  |  Interesting Insight |
|---------------|:-------------:|:-------------:|
| Sex Ed Mandated | Provided Educators with Strategies | There are 8 states in which sex ed is mandated but in which fewer than 70% of schools provide teachers strategies for teaching sex ed. In these states, teachers have a large amount of discretion. (However, it is unclear if this disparity between mandate and policy has an effect on student health. All of these 8 states have an STD index of 0.45 or lower, scoring better than about half of all states. In fact, one of the states, Vermont, has the lowest STD index of any state. By another metric, births per 1,000 girls, this result is less definitive: the states fall over a wide range, from 27.2 (Kentucky) to 8.8 (Vermont).) |"
cat(tablLP) # output the table in a format good for HTML/PDF/docx conversion
```

We cannot definitively conclude whether state mandates and teacher behavior are correlated, but there seems evidence that mandates leave much discretion to schools and teachers, and thus mandates do not correlate strongly with school and teacher practices. 

## Are school policies and teachers’ efforts effective in improving student health outcomes?

Because state mandates do not predict the comprehensiveness of school sex ed policies and teachers’ zeal in conveying the information to students, it may be that educator practices correspond to student health in a way that mandates do not. To investigate this idea, we’ve plotted school policies and teacher practices against health outcomes. Alternatively, users can choose to plot a metric related to student behavior to investigate whether behaviors are a better predictor of health outcomes than education metrics. 

<iframe height="600" width="100%" frameborder="no" title="SchoolScatter" src="https://anayak23.shinyapps.io/school_edu_scatter/"> </iframe>


```{r tableScatter, echo=FALSE, message=FALSE, warnings=FALSE, results='asis'}
##Corresponds with school_edu_scatter
tablScatter <- "
| Education Mandate        | Resource Offered by High Schools  |  Interesting Insight |
|---------------|:-------------:|:-------------:|
| Increased Student Knowledge on STD Prevention | STD Index  | This display provides some evidence that teachers’ efforts to improve students' knowledge on STD prevention improves their ability to protect themselves from STDs. |
| Increased Student knowledge on HIV Prevention | HIV Per 100K | Similarly, this display tells an optimistic story about teachers’ ability to help students protect themselves from HIV. |"
cat(tablScatter) # output the table in a format good for HTML/PDF/docx conversion
```

# What can spacial data show us about potential regional differences in mandates and outcomes?

<iframe height="1500" width="100%" frameborder="no" title="Spacial" src=" https://csarosi.shinyapps.io/spacial_data/"> </iframe>

```{r tableSpacial, echo=FALSE, message=FALSE, warnings=FALSE, results='asis'}
##Corresponds with spacial_data
tablSpacial <- "
| Education Mandate        | Resource Offered by High Schools  |  Interesting Insight |
|---------------|:-------------:|:-------------:|
| Increased Student Knowledge on STD Prevention | STD Index  | This display provides some evidence that teachers’ efforts to improve students' knowledge on STD prevention improves their ability to protect themselves from STDs. |
| Increased Student knowledge on HIV Prevention | HIV Per 100K | Similarly, this display tells an optimistic story about teachers’ ability to help students protect themselves from HIV. |"
cat(tablSpacial) # output the table in a format good for HTML/PDF/docx conversion
```

# Conclusions

Before we can draw any conclusions from our data and visualizations, we must first acknowledge the limitations of this data. 

The first is that, because we drew different variables from different sources, some data points don’t reflect data from the same years. While some sources are from an ambiguous time period, most are from the years 2017-2019. Presumably, our variables haven’t changed in a drastic way in this short period, though they could have.  

Another limitation is that our metrics of school practices and teacher attitudes relied on self-reporting from the state. States seem to differ in their infrastructures for collecting this information; for example Colorado presented wide ranges as estimates, of which we decided to take the mean. The disparity in states’ care and caution in collecting information is likely not independent from their mandates, policies, and practices regarding sex ed, and for this reason this could be a significant limitation.  

We  can make a few primary conclusions responsive to our research questions.

1. Statewide sexual health  education mandates do not seem to correlate with student’s health outcomes.

2. We cannot definitively say if there is a causal effect, but we think Conclusion 1 may be true in part because sexual health education mandates do not exactly predict what actually occurs in schools in particular states. School policies and teachers’ desire to emphasize the material seem to be independent from mandates. 

3. While strict mandates don’t correspond to better health outcomes, there is strong evidence that teachers’ buy-in to the material do correspond to better health outcomes. In particular, there seems to be a strong effect of teacher practices regarding STD and HIV prevention in decreasing STD and HIV rates (though we cannot assume this is a causal relationship.) On the other hand, schools that provide sex ed teachers the materials they need, going beyond state mandate guidelines, have not seen the reward of better student health outcomes. 

We would recommend the following to state legislatures, schools, or educators looking to improve the student health outcomes in their state. 

1. Initiatives that increase teacher buy-in, confidence, and comfort with the material will have the most significant impact of student health outcomes. It might be useful to put mediated groups together where teachers can talk about these topics and how they feel teaching them, as well as share their own techniques with others. 

2. A focus on the efficacy of the strategies that are recommended to teachers has the potential to go a long way, since states in which these strategies are most likely to be implemented also see high STD infection rates. 


# References

Alarms.org. “Sexually Transmitted Diseases by State.” National Council For Home Safety and Security, 17 Feb. 2020, [www.alarms.org/std-statistics/](www.alarms.org/std-statistics/).   

Guttmacher Institute. “Sex and HIV Education.” Guttmacher Institute, 2 Nov. 2020, [www.guttmacher.org/state-policy/explore/sex-and-hiv-education](www.guttmacher.org/state-policy/explore/sex-and-hiv-education).   

“Teen Birth Rate Comparison, 2018: Power to Decide.” Teen Birth Rate Comparison, 2018 | Power to Decide, 2020, [powertodecide.org/what-we-do/information/national-state-data/teen-birth-rate](powertodecide.org/what-we-do/information/national-state-data/teen-birth-rate).   

U.S. Department of Health & Human Services. “State School-Based Health Education Law Summary Reports.” Centers for Disease Control and Prevention, Centers for Disease Control and Prevention, 20 June 2018, [www.cdc.gov/healthyyouth/policy/state_law_summary_reports.html](www.cdc.gov/healthyyouth/policy/state_law_summary_reports.html). 